Gradual Soundness: Lessons from Static Python

### 2022-01-04 pj-7.1/paper.tex
pj: 3, quick draft of technical figures
idea is, use a very simple model to show where dyn comes in:

- basic surface lang. with function defs, class defs, and chkdict
- evaluation types show that Union and PyDict become Dyn
- grammar disallows nested classes
- typing judgment shows class override restrictions
- typing judgment shows pydict shallowness, chkdict depth
- subtyping / compatibility = upcast, materialization = downcast

the materialize rules need run-time casts,
we need to show that the casts are simple & full,
and argue that the cast locations are a subset of Nom

soundness argume

\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

\newcommand{\spexpr}{e}
\newcommand{\spvalue}{v}

@@

%\usepackage{cleveref}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
(\code{Union[T, None]}, or \code{Optional[t0]}).

@@

\section{Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

@@

%% TODO 2021-12-28
%% - model soundness argument = similar to Nom, Redex tests
%% - while True break another example of modeling helping out
%%   it's a second implementation, helpful to find bugs
%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us

@@

\subsection{Surface Language}

\begin{figure}[t]
  %% ported types from redex model
  \begin{langarray}
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptinstanceof{\sptclass} \langmid
      \sptfun{\sptype}{\sptype} \langmid
      \sptunion{\sptclass}{\sptclass}
  \\
    \sptclass & \defeq &
      \sptint \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptpydict{\sptclass}{\sptclass} \langmid
      \sptchkdict{\sptclass}{\sptclass} \langmid
      \mbox{(user defined class)}
  \end{langarray}

  \begin{center}\parbox{0.8\columnwidth}{
    Abbreviation: $\sptoptional{\sptclass} \defeq \sptunion{\sptclass}{\sptnone}$

    EXCEPT THAT $\sptoptional{\sptclass}$ may be a generic parameter just like any other class $\sptclass$

    Function types cannot expect or return other function types.
  }\end{center}

  \caption{Surface Types}
  \label{f:surface-types}
\end{figure}

\subsection{Evaluation Language, Optimizations}

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

  \caption{Surface Types to Evaluation Types}
  \label{f:surface-to-eval-types}
\end{figure}

@@

@@

@@

@@

>>>>

\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~#1(#2)\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[1]{\code{chkdict}(#1)}
\newcommand{\spapp}[2]{#1(#2)}
\newcommand{\spref}[2]{#1[#2]}
\newcommand{\spset}[3]{#1[#2] = #3}

@@

%\usepackage{cleveref}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).

@@

\section{Model of Core \SP{}}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

%% TODO 2021-12-30 outline
%% NOTE strict modules not modeled
%% - what is language
%% - what properties
%% - what did in model, how to navigate?
%% - bug reports

@@

%% TODO 2021-12-28
%% - model soundness argument = similar to Nom, Redex tests
%%   - Nom has dispatch modes .. SP does not, simpler
%%     - are dyn -> stat overrides allowed and do they affect soundness?
%% - while True break another example of modeling helping out
%%   it's a second implementation, helpful to find bugs
%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us

@@

%% TODO model omits primitives because their typing is easy, separate hierarchy, clear boundaries
%% TODO paper-model focus on boundaries ... where can dyn slip in ... not a full semantics ...

\subsection{Surface Syntax and Types}

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr} \ldots}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr} \ldots}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
  \\ & &
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spref{\spexpr}{\spexpr} \langmid
      \spset{\spexpr}{\spexpr}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptobject \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype} \ldots}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype} \ldots},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

%%  \begin{center}\parbox{0.8\columnwidth}{
  \bigskip
    \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}
%%  }\end{center}

  \caption{Surface Syntax and Types}
  \label{f:surface-types}
\end{figure}

%% TODO assume well-formed / normal-form
%% - no circular field defs (x:t = x)
%% - unions are tidy ... Opt[Dyn] = Dyn
%% kk

%% TODO surface typing judgment omitted
%% - straightforward
%% - eval typing far more interesting
%% - for now, main goal is to narrow unions ... that's kinda boring / not worth effort reading & writing

\subsection{Evaluation Types, Materialization, Casts}

\begin{figure}[t]
  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
      \\\\
      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
    }{
      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
    }
  \end{mathpar}

  \fbox{$\sptenv \wtprog \spstmt \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
    }

    \inferrule*{
      \sptenv_1 = \sptfundef{\spf_0}{\sptype_0}{\sptype_1},~\sptenv_0
      \\\\
      \sptenv_1 \wtexpr \spexpr_0 : \sptype_1
    }{
      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\sptype_0}}{\sptype_1}{\spexpr_0} \dashv \sptenv_1
    }

    \inferrule*{
      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0} \ldots}{\sptfundef{\spf_n}{\spc_0, \sptype_{dn}}{\sptype_{cn}}}
      \\\\
      %% TODO
      \emph{lookup parent}
      \\\\
      \emph{overrides are subtypes (not merely compatible)}
      \\\\
      \emph{fields and methods are well typed}
    }{
      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0} \ldots}{\spfundef{\spf_n}{\spself, \spann{\spx_n}{\sptype_{dn}}}{\sptype_{cn}}{\spexpr_n}} \dashv \sptenv_1
    }
  \end{mathpar}

  \fbox{$\sptenv \wtexpr \spexpr : \sptype$} (selected rules)
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spcompat \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spmatr \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

    \inferrule*{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*{
      \sptfundef{\spf_0}{\sptype_0}{\sptype_1} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptype_1
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_1
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptype_1
    }

  \end{mathpar}

  \caption{Evaluation Typing Judgment}
  \label{f:eval-typing}
\end{figure}

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

  \fbox{$\sptype \spsubt \sptype$}
  \begin{mathpar}
    \inferrule*{
    }{
      \sptype_0 \spsubt \sptobject
    }

    \inferrule*{
    }{
      \sptbool \spsubt \sptint
    }

    \inferrule*{
      \sptype_0 \spsubt \sptype_1
    }{
      \sptype_0 \spsubt \sptoptional{\sptype_1}
    }

    \inferrule*{
    }{
      \sptype_0 \spsubt \sptype_0
    }

    \inferrule*{
      \sptype_0 \spsubt \sptype_1
      \\
      \sptype_1 \spsubt \sptype_2
    }{
      \sptype_0 \spsubt \sptype_2
    }
  \end{mathpar}

  \fbox{$\sptype \spcompat \sptype$}
  \begin{mathpar}
    \inferrule*{
      \sptype_0 \spsubt \sptype_1
    }{
      \sptype_0 \spcompat \sptype_1
    }

    \inferrule*{
    }{
      \sptype_0 \spcompat \sptdyn
    }
  \end{mathpar}

  \fbox{$\sptype \spmatr \sptype$}
  \begin{mathpar}
    \inferrule*{
    }{
      \sptdyn \spmatr \sptype_0
    }
  \end{mathpar}

  \caption{TBA Surface Types to Evaluation Types}
  \label{f:surface-to-eval-types}
\end{figure}

@@

\subsubsection{Informal Connection to Nom}

\SP{} types restrict the use of Nom,
thus model is kind of a restriction of Nom.

So since Nom is sound for any sort of dyn mixing,
and SP allows fewer mixings,
and both Nom and SP do similar checks at boundaries,
can begin to see how SP could be sound.

Or perhaps mention this at the start.
SP similar to Nom, similar checks, keep that in mind.

\subsection{Redex Development, Navigation, Integration, Conformance Tests}

FILL

@@

%% TODO unions tracked statically in SP, erased to dyn unless narrowed

@@

\subsection{Python Interaction}

FILL

extend model with Python ... two languages ... boundaries ... argue for soundness

@@

%% 2021-12-30 Whence Nominal GT?
%%
%% KC: I think in general adding gradual typing to Java/C# (an also Python) is boring:
%% - MonNom allows you to cast between an arbitrary type and dyn.
%% - dyn is effectively Object
%% - Upcasts, casts from an arbitrary type to Object, are already in the language
%% - Downcasts, casts from Object to an arbitrary type is also possible, but must be explicit in the fully static language
%% So MonNom effectively gives programmers another way to name Object, with which downcasts can be automatically inserted by the compiler.
%% I think this sentence covers all the interesting stuff about MonNom.
%%
%% SK: This matches my intuition that gradual nominal is just not very
%% interesting. KC puts this on a technical footing with his very astute comment,
%% which to me can be summarized as "every nominal language that has Object is
%% already gradual" [leaving aside the boring non-Object base type distinction].
       So the entire remaining game is one of making the programming interface
%% as painless as possible. Some of that is just a trivial matter of
%% syntactic defaults; some of it is making the type-checker more forgiving
%% and replacing static checks with dynamic checks. There doesn't seem to
%% be much else left.
